---
title: Counting barcodes
author:
- name: Aaron Lun
  affiliation: Genentech gRED B&CB
  email: luna@gene.com
output:
  BiocStyle::html_document:
    toc_float: yes
package: gp.sa.screen
vignette: >
  %\VignetteIndexEntry{1. Counting barcodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE)
library(BiocStyle)
library(gp.sa.core)
```

# Overview

The `r GRANpkg("gp.sa.screen")` package implements several methods to quantify the frequency of barcodes in a screen sequencing experiment.
These operate on the raw FASTQ files, yielding a matrix of counts for each barcode in each sample for downstream analysis.
Quantification can be performed for both single and combinatorial barcodes.
Only single-end sequencing data are currently supported.

# Counting single barcodes

The "barcode" is considered to be the entire sequence structure that is used to label a particular molecular or biological entity.
This obviously includes regions of variable sequence that distinguish one barcode from another,
as well as constant sequences that flank or separate the variable regions.
The most common barcode design uses only one variable region, which we refer to as a "single barcode".

```
# Example of a single barcode:

CAGCAGCATGCTGNNNNNNNNNNNNNNCAGCATCGTGC
-------------^^^^^^^^^^^^^^-----------
constant     variable      constant
flank        region        flank
```

To demonstrate, let's mock up a FASTQ file for some barcode data.
For simplicity, we will assume that each read only contains the barcode.
However, it is entirely permissible to have additional (unknown) flanking sequences in the read.

```{r}
# Our pool of known variable sequences, one per barcode:
known <- c("AAAAAAAA", "CCCCCCCC", "GGGGGGGG", "TTTTTTTT")

# Mocking up some sequence data, where each read randomly contains 
# one of the variable sequences, flanked by constant regions. 
library(Biostrings)
chosen <- sample(known, 1000, replace=TRUE)
reads <- sprintf("GTAC%sCATG", chosen)
names(reads) <- sprintf("READ_%i", seq_along(reads))

# Writing to a FASTQ file.
single.fq <- tempfile(fileext=".fastq")
writeXStringSet(DNAStringSet(reads), file=single.fq, format="fastq")
```

We quantify single barcodes across one or more files using the `matrixOfSingleBarcodes()` function.
This produces a matrix of frequencies of each barcode (row) in each file (column).
The order of rows corresponds to the order of variable regions in `known` that distinguishes each barcode.

```{r}
library(gp.sa.screen)
out <- matrixOfSingleBarcodes(single.fq, 
    flank5="GTAC", flank3="CATG", choices=known)
out
```

We specify the constant sequences that immediately flank the variable region^[A template structure can also be specified in `template=`, which may be more convenient.].
This improves stringency of barcode identification, as the barcode will not be recognized in the read sequence unless the constant sequences match perfectly.
It also improves speed as the matching of the variable sequence is only performed at positions along the read where the flanking constant sequences have matched.

Obviously, users should not supply the full length of the flanking sequence if that sequence is not captured in the read.
For example, the 3' flanking sequence may be lost in experiments using short reads.
In such cases, an empty string should be specified as `flank3=`.

# Counting combinatorial barcodes

A more complex experimental design involves combinatorial barcodes where multiple variable regions are randomly ligated to form a single sequence.
This exploits combinatorial complexity to generate a large number of unique barcodes from a limited pool of known variable sequences.

```
# Example of a combinatorial barcode:

CAGCTANNNNNNNNNNCACGNNNNNNNNNNCAGCT
------^^^^^^^^^^----^^^^^^^^^^-----
      variable      variable
```

To demonstrate, let's mock up another FASTQ file of single-end read data.
Again, for simplicity, we will assume that each read sequence contains only the barcode.

```{r}
# Our pool of known variable sequences:
known <- c(
    "AAAA",
    "CCCC",
    "GGGG",
    "TTTT"    
)

# Mocking up some sequence data, where each read randomly contains 
# two of the variable sequences within a template structure.
library(Biostrings)
chosen1 <- sample(known, 1000, replace=TRUE)
chosen2 <- sample(known, 1000, replace=TRUE)
reads <- sprintf("GTAC%sCATG%sGTAC", chosen1, chosen2)
names(reads) <- sprintf("READ_%i", seq_along(reads))

# Writing to a FASTQ file.
combo.fq <- tempfile(fileext=".fastq")
writeXStringSet(DNAStringSet(reads), file=combo.fq, format="fastq")
```

We quantify combinatorial barcodes across one or more files using the `matrixOfComboBarcodes()` function.
This requires a template for the barcode structure to specify how the variable sequences are used to construct the final barcode.
It is permissible for each variable sequence to be sampled from a different known pool, though in this case, both regions are sampled from the same pool.

```{r}
out <- matrixOfComboBarcodes(combo.fq,
    template="GTACNNNNCATGNNNNGTAC",
    choices=list(known, known))    
out$combination
out$counts
```

This function produces a matrix of frequencies of each barcode (row) in each file (column).
The identity of the variable regions in each combinatorial barcode is contained in the `combination` data frame, which contains integer indices to the variable sequences in `known.barcodes`.

# Further options

## Supporting mismatches and deletions

Matching allows for _no more than_ one substitution or deletion within any variable region.
If multiple variable regions are present in a combinatorial barcode, only one substitution or deletion can be present across all variable regions.
This is equivalent to an edit distance threshold of one (across the variable regions).

Allowing for mismatches or deletions improves recovery of the barcode in the presence of sequencing errors.
While such errors are rare per base^[At least on Illumina machines.], they will be more likely to occur when considering the entire length of the barcode.
Substitutions are most common, followed by deletions; insertions are extremely rare and are not supported.

Mismatch tolerance is turned on by default in all counting functions.
Users can disable these selectively by setting `substitutions=FALSE` or `deletions=FALSE`.
This may be necessary to achieve high stringency for short barcodes.

## Parallelization across files

Counting functions are parallelized across files using the `r Biocpkg("BiocParallel")` framework.
Each file is processed separately on a single worker to create the final matrix of counts.

```{r}
# Pretend that these are different samples:
all.files <- c(single.fq, single.fq, single.fq)

# Parallel execution:
library(BiocParallel)
out <- matrixOfSingleBarcodes(all.files, 
    flank5="GTAC", flank3="CATG", choices=known,
    BPPARAM=SnowParam(2))
out
```

Users are advised to read the `r Biocpkg("BiocParallel", "Introduction_To_BiocParallel.pdf", "relevant documentation")` to determine the most effective parallelization approach for their system.
For example, users working on a cluster should use `BatchJobsParam()`, while users on non-Windows machines may prefer to use `MulticoreParam()`.

# Session information

```{r}
sessionInfo()
```
